---
title: "Linear Algebra in R with Applications to Linear Regression"
author: "Patrick Walls"
date: "June 23, 2016"
output: pdf_document
---

1. Vectors
2. Matrices
3. Matrix operations: multiplication, inverse and transpose
4. Linear regression

## 1. Vectors

We can create vectors in several ways using the following functions:

* `c(value,value, ... ,value)` concatenates its arguments to construct a vector
* `seq(from,to,by=1,length)` creates a vector with sequential values
* `from:to` (colon operator) is a shorthand method for `seq(from,to,by=1)`
* `rep(x,times)` takes a value/vector and repeats it according to the parameter _times_

For example:

```{r}
c(1,0,-1,2)
seq(0,10,by=2)
0:10
rep(0,times=5)
```

We can also create vectors with random entries using the functions:

* `runif(n,min=0,max=1)` generates a vector of length _n_ sampled uniformly from the interval [_min_,_max_]
* `rnorm(n,mean=0,sd=1)` generates a vectors of length _n_ sampled from the normal distribution
* `sample(x,n,replace=FALSE)` generates a vector of length _n_ sampled from the vector _x_ (with or without replacement)

For example:

```{r}
runif(3,0,5)
rnorm(4)
sample(0:6,20,replace=TRUE)
```

Note that arithmetic operations are performed elementwise on vectors:

For example:

```{r}
1:10 / rep(2,10)
1:5 * 1:5
```

### Example: Simulating Random Data

* Create a vector _x_ of length 50 sampled uniformly from the interval [0,10]
* Create a vector _err_ of length 50 sampled from the standard normal distribution
* Create a vector _y_ by the formula _y_ = 2 _x_ + 1 + _err_
* Plot _x_ versus _y_ in a scatter plot
* Create vectors _X_ = [0,1,2,...,10] and _Y_ = 2 _X_ + 1 and plot the line _X_ versus _Y_ in the same plot

```{r}
n <- 50
x <- runif(n,0,10)
err <- rnorm(n)
y <- 2*x + 1 + err
plot(x,y)
X <- 0:10
Y <- 2*X + 1
lines(X,Y,'l')
title('Simulating Random Data')
```

### Exercise

Create the following vectors without using the function `c()`:

* $[ 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3]$
* $[ 5, 4, 3, 2, 1]$
* $[ -100, -75, -50, -25, 0, 25, 50, 75, 100]$

### Exercise

1. Plot the function $y = e^{-x^2}$ for $x \in [-5,5]$.
2. Plot 10,000 random points $(x,y)$ where $x$ is sampled from the uniform distribution and $y$ is sampled from the standard normal distribution.

```{r}
x <- runif(10000)
y <- rnorm(10000)
plot(x,y,col=rgb(0,0,0,0.2),pch=16,cex=0.5)
```

## 2. Matrices

We use the function `matrix(data,nrow,ncol,byrow=FALSE)` to create a matrix from the entries in _data_. The shape of the matrix is determined by _nrow_ and _ncol_. The input _data_ can be a vector or a value and the `matrix()` function will fill in the values of the matrix. For example:

```{r}
matrix(1:4,2,byrow=TRUE)
matrix(1,3,2)
matrix(sample(-9:9,9,replace=TRUE),nrow=3)
```

Arithmetic operations are **performed elementwise**. For example:

```{r}
A <- matrix(1:4,2,2,byrow=TRUE)
A * A
```

```{r}
B <- matrix(2,2,2,byrow=TRUE)
A / B
```

### Exercise

1. Create the following matrices:

* $\begin{bmatrix} 2 & 3 \\ -4 & 1 \end{bmatrix}$
* $\begin{bmatrix} 1 & 1 & 0 \\ -2 & 5 & 2 \end{bmatrix}$

2. Create a 5 by 5 matrix with alternating entries 1, -1, 1, -1, etc. (starting in position (1,1) and proceeding by row).

3. Create a 10 by 10 matrix with entries randomly sampled from $\{0,1\}$.

4. Create the 4 by 4 identity matrix.

## 3. Matrix Operations: Multiplication, Inverse and Transpose

We must use the `%*%` operator to perform matrix multiplication. For example:

$$
\begin{bmatrix} 2 & 3 \\ -1 & 1 \end{bmatrix}
\begin{bmatrix} -1 & 2 \\ 0 & 2 \end{bmatrix} =
\begin{bmatrix} -2 & 10 \\ 1 & 0 \end{bmatrix}
$$

```{r}
A <- matrix(c(2,3,-1,1),nrow=2,byrow=TRUE)
A
B <- matrix(c(-1,2,0,2),nrow=2,byrow=TRUE)
B
A %*% B
```

We use the function `t()` to compute the transpose of a matrix.

```{r}
A <- matrix( sample(-1:1,16,replace=TRUE) , nrow=4 , byrow=TRUE )
A
t(A)
```

For example, we can take the dot product of a vector _v_ with itself using the operator `%*%` along with the transpose function _t(v)_:


```{r}
v <- matrix( c(1,2) , nrow=1 )
w <- t(v)
v %*% w # 1^2 + 2^2 = 5
```

We use the function `solve()` to compute the inverse of a matrix:

```{r}
A <- matrix( sample(-10:10,16,replace=TRUE) , nrow=4 , byrow=TRUE )
A
B <- solve(A)
B
A %*% B
```

### Exercise

1. Create a 3 by 3 matrix $A$ with integer entries randomly sampled from $[-99,99]$ and compute $A^T A$. Do you notice something special about the result?

2. Find the inverse of $\begin{bmatrix} 1 & 2 \\ 3 & 5 \end{bmatrix}$ and confirm $A^{-1} A = I$.

## 4. Linear Regression

Consider a set of $n$ data points: $(x_1,y_1) , \dots , (x_n,y_n)$. The linear model which best fits the data (by minimizing the sum of squared errors) is:

$$
y = a_1 + a_2 x
$$

where the coefficients $a_1$ and $a_2$ are determined by

$$
A = ( M^T M )^{-1} M^T Y
$$

where

$$
A = \begin{bmatrix} a_1 \\ a_2 \end{bmatrix} \ , \ M = \begin{bmatrix} 1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots \\ 1 & x_n \end{bmatrix} \ , \ Y = \begin{bmatrix} y_1 \\ \vdots \\ y_n \end{bmatrix}
$$

### Exercise

Write a function `linreg(x,y)` which takes vectors x and y of the same length and return the coefficients of the linear regression. Also, the function should plot the points and the linear model, and return the R2 value.

```{r}
linreg <- function(x,y) {
  M <- matrix(nrow=length(x),ncol=2)
  M[,1] <- 1 # Set all entries in first column to 1
  M[,2] <- x # Set the second column to the x values
  Y <- matrix(y,ncol=1)
  A <- solve( t(M) %*% M ) %*% t(M) %*% Y
  
  plot(x,y,col=rgb(0,0,0,0.5),pch=16)
  X <- seq(min(x),max(x),length=3)
  lines(X, A[1,1] + A[2,1]*X,col='red',lwd=5)
  
  ymean <- sum(y) / length(y)
  total_err <- sum( (y - ymean)^2 )
  residues <- sum( (M %*% A - y)^2 )
  R2 <- 1 - residues / total_err
  
  print( paste('Linear model: y = ',
               round(A[1,1],2), ' + ',
               round(A[2,1],2),
               'x, and R2 = ',
               round(R2,2),'.',sep=''))
  return(A)
}
```

```{r}
x <- runif(1000,0,5)
err <- rnorm(1000)
y <- 2.7*x + 1.4 + err
coefs <- linreg(x,y)
```
